---
title: "CourseAssignment#1"
author: "Eric Kruep"
date: "Saturday, April 18, 2015"
output: html_document
---

###Loading and preprocessing the data

Step 1: Download data from provided link to local file (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

Step 2: Load the data from local file into R

```{r}
data <- read.csv("./data/activity.csv")

summary(data)
head(data)
str(data)

```

Step 3: Tidy the data for analysis - Convert the data field to Date class and interval field as factor class

```{r}
data$date <- as.Date(data$date, format = "%Y-%m-%d")
data$interval <- as.factor(data$interval)
```


###What is mean total number of steps taken per day?

Step 1: Calculate the total number of steps taken per day

```{r}
totalSteps <- aggregate(steps ~ date, data, sum, na.rm=TRUE)
head(totalSteps)

```

Step 2: Make a histogram of the total number of steps taken each day
```{r}
library(ggplot2)

ggplot(totalSteps, aes(x = steps)) + 
       geom_histogram(fill = "lightblue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of Times in a Day (Count)") + theme_bw() 


```

Step 3: Calculate and report the mean and median of the total number of steps taken per day
```{r}
meanSteps <- mean(totalSteps$steps, na.rm=TRUE)
medianSteps <- median(totalSteps$steps, na.rm=TRUE)

```

- The **mean** number of steps taken per day is **`r meanSteps`**.
- The **median** number of steps taken per day is **`r medianSteps`**. 



###What is the average daily activity pattern?

Step 1: Make a time series plot (i.e., type = "1") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

We find the aggregation of steps by intervals of 5-minutes and then convert these intervales to integers. We save them as a data.frame

```{r}
stepsPerInterval <- aggregate(data$steps, 
                                by = list(interval = data$interval),
                                FUN=mean, na.rm=TRUE)

## Convert to integers to aid plotting
stepsPerInterval$interval <- as.integer(levels(stepsPerInterval$interval)[stepsPerInterval$interval])
colnames(stepsPerInterval) <- c("interval", "steps")

```

Then we can plot the data.frame
```{r}
ggplot(stepsPerInterval, aes(x = interval, y = steps)) +   
        geom_line(color="green", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of Steps") +  
        theme_bw()

```

Step 2: Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
maxInterval <- stepsPerInterval[which.max(  
        stepsPerInterval$steps),]

maxInterval

```
The **`r maxInterval$interval`th** interval has the maximum value of **`r maxInterval$steps`** steps.

##Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as 'r NA'). The presence of missing days may introduce bias into some calculations or summaries of the data.

Step 1: Calculate and report the total number of missing values in the dataset (i.e., the total number of rows with `r NA`s)

```{r}
missingValues <- sum(is.na(data$steps))
```
The total number of missing values is **`r missingValues`**.

Step 2: Devise a strategy for filling in all of the missing values in the dataset. The strategy chosen here is the following: replace them with the mean value at the same interval across days

```{r}
naFill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

dataFill <- data.frame(  
        steps = naFill(data, stepsPerInterval),  
        date = data$date,  
        interval = data$interval)
str(dataFill)

```

Check for any remaining missing values

```{r}
sum(is.na(dataFill$steps))

```

Step 3: Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r}
fillStepsPerDay <- aggregate(steps ~ date, dataFill, sum)
colnames(fillStepsPerDay) <- c("date","steps")
```


Step 4:  Make a histogram of the total number of steps taken each day, calculate and report the mean and median total number of steps taken per day.

Create the histogram with ggplot

```{r}
ggplot(fillStepsPerDay, aes(x = steps)) + 
       geom_histogram(fill = "lightblue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of Times in a Day (Count)") + theme_bw() 

```

Calculate and report the mean and median total number of steps taken per day. 

```{r}
meanFillStepsPerDay <- mean(fillStepsPerDay$steps, na.rm=TRUE)
medianFillStepsPerDay <- median(fillStepsPerDay$steps, na.rm=TRUE)

```

- The **mean** number of steps taken per day is **`r meanFillStepsPerDay`**.
- The **median** number of steps taken per day is **`r medianFillStepsPerDay`**. 

**CONCLUSION:** These values are actually quite similar to the values with un-imputed data. 

###Are there differences in activity patterns between weekdays and weekends?

Step 1: Use the weekdays() function with the filled-in missing values data to look at differences in between weekdays and weekends

Step 2: Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

Step 3: Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.


```{r}
weekdaysSteps <- function(data) {
    weekdaysSteps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=TRUE)
    ## Convert to integers for plotting
    weekdaysSteps$interval <- 
            as.integer(levels(weekdaysSteps$interval)[weekdaysSteps$interval])
    colnames(weekdaysSteps) <- c("interval", "steps")
    weekdaysSteps
}

dataByWeekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekendData <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekdayData <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekendSteps <- weekdaysSteps(weekendData)
    weekdaySteps <- weekdaysSteps(weekdayData)

    weekendSteps$dayofweek <- rep("weekend", nrow(weekendSteps))
    weekdaySteps$dayofweek <- rep("weekday", nrow(weekdaySteps))

    dataByWeekdays <- rbind(weekendSteps, weekdaySteps)
    dataByWeekdays$dayofweek <- as.factor(dataByWeekdays$dayofweek)
    dataByWeekdays
}

dataWeekdays <- dataByWeekdays(dataFill)

```


Here is a panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends:

```{r}
ggplot(dataWeekdays, aes(x = interval, y = steps)) + 
        geom_line(color="red") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of Steps") +
        theme_bw()

```

**CONCLUSION:** Activity on the weekends appears to be more consistently higher than during the weekdays. 






